---
title: "Malassezia Paper Script"
author: "Ronja Steinbach -- ronja36@hawaii.edu"
date: "Update 11/11/21"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r setup chunk, setup, include = FALSE, cache=FALSE, message=FALSE, warning=FALSE}
if (!require('knitr')) install.packages('knitr'); library('knitr')
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align='center')
########## Load packages
# if you don't have them already in your library, the following code can be used
# this will install the package if it does not already exist on your machine

# load packages
if (!require("pacman")) install.packages("pacman") # for rapid install if not in library

# use pacman to load all the packages you are missing!
pacman::p_load('knitr', 'lme4', 'lmerTest', 'effects', 'dplyr', 'plotrix', 'car', 'bookdown')

library(phyloseq); packageVersion("phyloseq") #version 1.34.0
library(Biostrings); packageVersion("Biostrings") #version 2.58.0
library(ggplot2); packageVersion("ggplot2") #version 3.3.3
library("plyr")
library("vegan") #2.5-7
library("decontam") 
library("rgdal")
library("ape")
library(ggtree)
library(treeio)
library(scales)
library(phyloseq)
library(scales)
library(viridis)
library(ggstance)
```
### Set your working directory and read in the Phyloseq object
```{r, echo=TRUE, results='hide'}
getwd() # get working directory
setwd("/Users/ronjasteinbach/Desktop/Data Analysis UROP/ForRonja/UROP-Project-Spring2021/Paper_Scripts/Mal_Paper_R_Materials") # set your working directory

#Read in the Phyloseq Object ----
physeq2 <- readRDS("Mala_Physeq_Object") #We are naming the phyloseq object we are working "physeq2"
```
### Subset so that sample sums > 10 and taxa > 0
```{r}
physeq2 #350 taxa, 473 samples

#Creating a physeq object of the more abundant ones by pruning physeq to sample sums greater than 10 and taxa sums greater than 0 (needed so there are no zeroes for ordinations)
physeq_abundant <- subset_samples(physeq2, sample_sums(physeq2)>10) #sample sums greater than 10
physeq_abundant <- prune_taxa(taxa_sums(physeq_abundant) > 0, physeq_abundant) # sample sums > 10 and taxa > 0
any(taxa_sums(physeq_abundant) == 0) #make sure all taxa sums are greater than 0
physeq_abundant #279 taxa, 128 samples
#saveRDS(physeq_abundant, "Mala_Physeq_Abundant") #Save file with only the abundant data
physeq_abundant <- subset_samples(physeq_abundant, location_broad!= "Lab") #remove the data that came from lab samples and not Waimea
physeq_abundant #279 taxa, 127 samples
#write.csv(otu_table(physeq_abundant), file="279_taxa_otu_table.csv") #save OTU table
```

### Ordinations -- Bray Curtis Principal Coordinates Analysis (PCoA)
Next we are going to make PCoA ordinations based on EMPO Level 3 Category and Location
```{r, echo=TRUE, results='hide'}
# Transform data to proportions as appropriate for Bray-Curtis distances by finding relative abundance/normalizing the data
physeq_prop <- transform_sample_counts(physeq_abundant, function(OTU) OTU/sum(OTU))
max(otu_table(physeq_prop)) #1
sum(otu_table(physeq_prop)) #127 

ord.physeq <- ordinate(physeq_prop, method="PCoA", distance="bray") #Ordinate using the Bray distance matrix

# Plot the ordinations with multivariate t distributions and 95% CI Ellipses

#----Color = EMPO Level 3 Category, Shape = Location
ordplot <- plot_ordination(physeq_prop, ord.physeq, shape = "level3", color="location_broad", title="PCoA") #different shape for Level3 category, different color for location
ordplot <- ordplot + stat_ellipse(type = "t") + scale_colour_grey() #add 95% CI Ellipses and change to grayscale
print(ordplot)

ordplot_color <- plot_ordination(physeq_prop, ord.physeq, shape = "level3", color="location_broad", title="PCoA") #different shape for Level3 category, different color for location'
ordplot_color <- ordplot_color + stat_ellipse(type = "t")  #add 95% CI Ellipses
print(ordplot_color)

#---- Color and Shape = EMPO Level 3 Category
ordplot_level3 <- plot_ordination(physeq_prop, ord.physeq, shape = "level3", color="level3", title="PCoA") #different shape for Level3 category, different color for location
ordplot_level3 <- ordplot_level3 + stat_ellipse(type = "t") + scale_colour_grey() #add 95% CI Ellipses and change to grayscale
print(ordplot_level3)

ordplot_level3_color <- plot_ordination(physeq_prop, ord.physeq, shape = "level3", color= "level3", title="PCoA")  #different shape for Level3 category, different color for location'
ordplot_level3_color <- ordplot_level3_color + stat_ellipse(type = "t") #add 95% CI Ellipses 
print(ordplot_level3_color)

# ---- Color and Shape = Location
ordplot_location <- plot_ordination(physeq_prop, ord.physeq, shape = "location_broad", color="location_broad", title="PCoA") #different shape for location category, different color for location
ordplot_location <- ordplot_location + stat_ellipse(type = "t") + scale_colour_grey() #add 95% CI Ellipses and change to grayscale
print(ordplot_location)

ordplot_location_color <- plot_ordination(physeq_prop, ord.physeq, shape = "location_broad", color="location_broad", title="PCoA") #different shape for location category, different color for location
ordplot_location_color <- ordplot_location_color + stat_ellipse(type = "t") #add 95% CI Ellipses
print(ordplot_location_color)

#Save ordinations ---
#pdf("ordplotlevel3.pdf")
#ordplot_level3
#dev.off()

#pdf("ordplotlevel3_color.pdf")
#ordplot_level3_color
#dev.off()

#pdf("ordplotlocation.pdf")
#ordplot_location
#dev.off()

#pdf("ordplotlocation_color.pdf")
#ordplot_location_color
#dev.off()


#my_colors <- c("red", "orange", "yellow", "green", "blue")
#names(my_colors) <- #levels(factor(c(levels(sample_data(physeq_prop)$level3))))
#my_scale <- scale_fill_manual(name = "level3", values = my_colors)
```
### PERMANOVA and Beta Dispersion
```{r, echo=TRUE, results='hide'}
# PERMANOVA ---------------------
physeq.df <- data.frame(sample_data(physeq_prop)) #make a data frame for the sample/metadata

physeq_dist_bray <- phyloseq::distance(physeq_prop, method="bray") #calculate the Bray distance matrix

adonis2(physeq_dist_bray ~ level3 * location_broad, data=physeq.df) # calculate dissimilarity


#Betadispersion/Post-Hoc Tests (Unnecessary - no signficant results from PERMANOVA) -------------------
level3_factor <- factor(sample_data(physeq_prop)$level3)
exo_disper <- betadisper(physeq_dist_bray, level3_factor)
exo_disper

# Plots (not helpful) ---
#plot(exo_disper)
#boxplot(exo_disper, ylab = "Distance to centroid", xlab="Sample Type")
```

```{r}
physeq_abundant #279 taxa, 127 samples
melt_abund <- psmelt(physeq_abundant)
A <- subset(melt_abund, Letter=="A") #6604
B <- subset(melt_abund, Letter=="B") #0
C <- subset(melt_abund, Letter=="C") #0
D <- subset(melt_abund, Letter=="D")  #4064
E <- subset(melt_abund, Letter=="E") #0
F <- subset(melt_abund, Letter=="F") #0
G <- subset(melt_abund, Letter=="G") #889
H <- subset(melt_abund, Letter=="H") #0
I <- subset(melt_abund, Letter=="I") #2921
J <- subset(melt_abund, Letter=="J") #20701
K <- subset(melt_abund, Letter=="K") #127
L <- subset(melt_abund, Letter=="L") #127

Asum <- sum(A$Abundance) #239591
Dsum <- sum(D$Abundance) #34312
Gsum <- sum(G$Abundance) #24568
Isum <- sum(I$Abundance) #338
Jsum <- sum(J$Abundance) #1662639
Ksum <- sum(K$Abundance) #22
Lsum <- sum(L$Abundance) #44

Aanimal <- subset(A, level3=="Animal_corpus") #1976
Aplant <- subset(A, level3=="Plant_corpus") #1820
Asurface <- subset(A, level3=="Surface") #1820
Asediment <- subset(A, level3=="Sediment") #364
Awater <- subset(A, level3=="Water") #624

Danimal <- subset(D, level3=="Animal_corpus") #1216
Dplant <- subset(D, level3=="Plant_corpus") #1120
Dsurface <- subset(D, level3=="Surface") #1120
Dsediment <- subset(D, level3=="Sediment") #224
Dwater <- subset(D, level3=="Water") #384

Ganimal <- subset(G, level3=="Animal_corpus") #266
Gplant <- subset(G, level3=="Plant_corpus") #245
Gsurface <- subset(G, level3=="Surface") #245
Gsediment <- subset(G, level3=="Sediment") #49
Gwater <- subset(G, level3=="Water") #84

Ianimal <- subset(I, level3=="Animal_corpus") #874
Iplant <- subset(I, level3=="Plant_corpus") #805
Isurface <- subset(I, level3=="Surface") #805
Isediment <- subset(I, level3=="Sediment") #161
Iwater <- subset(I, level3=="Water") #276

Janimal <- subset(J, level3=="Animal_corpus") #6194
Jplant <- subset(J, level3=="Plant_corpus") #5705
Jsurface <- subset(J, level3=="Surface") #5705
Jsediment <- subset(J, level3=="Sediment") #1141
Jwater <- subset(J, level3=="Water") #1956

Kanimal <- subset(K, level3=="Animal_corpus") #38
Kplant <- subset(K, level3=="Plant_corpus") #35
Ksurface <- subset(K, level3=="Surface") #35
Ksediment <- subset(K, level3=="Sediment") #7
Kwater <- subset(K, level3=="Water") #12

Lanimal <- subset(L, level3=="Animal_corpus") #38
Lplant <- subset(L, level3=="Plant_corpus") #35
Lsurface <- subset(L, level3=="Surface") #35
Lsediment <- subset(L, level3=="Sediment") #7
Lwater <- subset(L, level3=="Water") #12

#_________ADGIJKL Abundance per level 3 category
Aanimalsum <- sum(Aanimal$Abundance) #26205
Aplantsum <- sum(Aplant$Abundance) #194247
Asurfacesum <- sum(Asurface$Abundance) #52
Asedimentsum <- sum(Asediment$Abundance) #0
Awatersum <- sum(Awater$Abundance) #19087

Danimalsum <- sum(Danimal$Abundance) #15575
Dplantsum <- sum(Dplant$Abundance) #18735
Dsurfacesum <- sum(Dsurface$Abundance) #0
Dsedimentsum <- sum(Dsediment$Abundance) #0
Dwatersum <- sum(Dwater$Abundance) #2

Ganimalsum <- sum(Ganimal$Abundance) #24567
Gplantsum <- sum(Gplant$Abundance) #1
Gsurfacesum <- sum(Gsurface$Abundance) #0
Gsedimentsum <- sum(Gsediment$Abundance) #0
Gwatersum <- sum(Gwater$Abundance) #0

Ianimalsum <- sum(Ianimal$Abundance) #167
Iplantsum <- sum(Iplant$Abundance) #29
Isurfacesum <- sum(Isurface$Abundance) #94
Isedimentsum <- sum(Isediment$Abundance) #0
Iwatersum <- sum(Iwater$Abundance) #48

Janimalsum <- sum(Janimal$Abundance) #154291
Jplantsum <- sum(Jplant$Abundance) #205662
Jsurfacesum <- sum(Jsurface$Abundance) #1206494
Jsedimentsum <- sum(Jsediment$Abundance) #449
Jwatersum <- sum(Jwater$Abundance) #95743

Kanimalsum <- sum(Kanimal$Abundance) #0
Kplantsum <- sum(Kplant$Abundance) #22
Ksurfacesum <- sum(Ksurface$Abundance) #0
Ksedimentsum <- sum(Ksediment$Abundance) #0
Kwatersum <- sum(Kwater$Abundance) #0

Lanimalsum <- sum(Lanimal$Abundance) #0
Lplantsum <- sum(Lplant$Abundance) #44
Lsurfacesum <- sum(Lsurface$Abundance) #0
Lsedimentsum <- sum(Lsediment$Abundance) #0
Lwatersum <- sum(Lwater$Abundance) #0

#______________________
AOahu <- subset(A, location_broad=="Oahu") #5928
AWoods_Hole <- subset(A, location_broad=="Woods_Hole") #312
AWashington <- subset(A, location_broad=="Washington") #364

DOahu <- subset(D, location_broad=="Oahu") #3648
DWoods_Hole <- subset(D, location_broad=="Woods_Hole") #192
DWashington <- subset(D, location_broad=="Washington") #224

GOahu <- subset(G, location_broad=="Oahu") #798
GWoods_Hole <- subset(G, location_broad=="Woods_Hole") #42
GWashington <- subset(G, location_broad=="Washington") #49

IOahu <- subset(I, location_broad=="Oahu") #2622
IWoods_Hole <- subset(I, location_broad=="Woods_Hole") #138
IWashington <- subset(I, location_broad=="Washington") #161

JOahu <- subset(J, location_broad=="Oahu") #18582
JWoods_Hole <- subset(J, location_broad=="Woods_Hole") #978
JWashington <- subset(J, location_broad=="Washington") #1141

KOahu <- subset(K, location_broad=="Oahu") #114
KWoods_Hole <- subset(K, location_broad=="Woods_Hole") #6
KWashington <- subset(K, location_broad=="Washington") #7

LOahu <- subset(L, location_broad=="Oahu") #114
LWoods_Hole <- subset(L, location_broad=="Woods_Hole") #6
LWashington <- subset(L, location_broad=="Washington") #7

#_________ADGIJKL Abundance per State
AOahusum <- sum(AOahu$Abundance) #239554
AWoods_Holesum <- sum(AWoods_Hole$Abundance) #6
AWashingtonsum <- sum(AWashington$Abundance) #31

DOahusum <- sum(DOahu$Abundance) #34312
DWoods_Holesum <- sum(DWoods_Hole$Abundance) #0
DWashingtonsum <- sum(DWashington$Abundance) #0

GOahusum <- sum(GOahu$Abundance) #1
GWoods_Holesum <- sum(GWoods_Hole$Abundance) #0
GWashingtonsum <- sum(GWashington$Abundance) #24567

IOahusum <- sum(IOahu$Abundance) #154
IWoods_Holesum <- sum(IWoods_Hole$Abundance) #26
IWashingtonsum <- sum(IWashington$Abundance) #158

JOahusum <- sum(JOahu$Abundance) #1529389
JWoods_Holesum <- sum(JWoods_Hole$Abundance) #94339
JWashingtonsum <- sum(JWashington$Abundance) #38911

KOahusum <- sum(KOahu$Abundance) #22
KWoods_Holesum <- sum(KWoods_Hole$Abundance) #0
KWashingtonsum <- sum(KWashington$Abundance) #0

LOahusum <- sum(LOahu$Abundance) #0
LWoods_Holesum <- sum(LWoods_Hole$Abundance) #44
LWashingtonsum <- sum(LWashington$Abundance) #0
```
### Heatmap
```{r}
#Relative Abundance Heatmaps and Data Transformations

# Surface always = rock swab (the only type of surface sampling we did)
dat <- read.csv("rel_abund_L3.csv")
dat
#Cube Root of Relative Abundance Heatmap (sample type adds up to 1)
# Surface always = rock swab (the only type of surface sampling we did)
dat$cube_root_rel <- (dat$Relative.Abundance)^(1/3) # make a new column with the square root of the relative abundance
heatmap_cube_root_rel <- ggplot(dat, aes(Level3, Clade)) +
  geom_tile(aes(fill = cube_root_rel), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("heatmap_cube_root_rel.pdf")
heatmap_cube_root_rel
#dev.off()

#Raw Abundance Heatmap
# Surface always = rock swab (the only type of surface sampling we did)
raw_dat <- read.csv("Raw_Reads.csv")
raw_dat$log10_Raw <- log10(raw_dat$Raw_Reads)
raw_dat$log10_Raw[which(!is.finite(raw_dat$log10_Raw))] <- 0
heatmap_raw <- ggplot(raw_dat, aes(Level3, Clade)) +
  geom_tile(aes(fill = log10_Raw), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("Raw_log_heatmap.pdf")
heatmap_raw
#dev.off()
#-------------------------------------------

#--------
#Relative Abundance
heatmap <- ggplot(dat, aes(Level3, Clade)) +
  geom_tile(aes(fill = Relative.Abundance), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("heatmap_rel.pdf")
heatmap
#dev.off()
#--------
#Log of Relative Abundance Heatmap (sample type adds up to 1)
# Surface always = rock swab (the only type of surface sampling we did)

dat$log10_rel <- log10(dat$Relative.Abundance) # take the log of the relative abundance
dat$log10_rel[which(!is.finite(dat$log10_rel))] <- 0 #replace all infinite values with 0 (since log 0 = inf)
heatmap_log_rel <- ggplot(dat, aes(Level3, Clade)) +
  geom_tile(aes(fill = log10_rel), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("heatmap_log_rel.pdf")
heatmap_log_rel
#dev.off()
#--------
#Log Plus One of Relative Abundance Heatmap (sample type adds up to 1)
# Surface always = rock swab (the only type of surface sampling we did)
dat$rel_one <- dat$Relative.Abundance + 1 # make a new data frame with relative abundance plus one
dat$log_rel_one <- log10(dat$rel_one) # take the log of relative abundance plus one (so it's positive)
heatmap_log_rel_one <- ggplot(dat, aes(Level3, Clade)) +
  geom_tile(aes(fill = log_rel_one), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("heatmap_log_rel.pdf")
heatmap_log_rel_one
#dev.off()
#--------
#Hellinger of Relative Abundance Heatmap (sample type adds up to 1)
# Surface always = rock swab (the only type of surface sampling we did)
dat$sqrt_rel <- sqrt(dat$Relative.Abundance) # make a new column with the square root of the relative abundance
dat$Hellinger <- log10(dat$sqrt_rel) # make a new column with the log of the square root of relative abundance
heatmap_Hellinger <- ggplot(dat, aes(Level3, Clade)) +
  geom_tile(aes(fill = Hellinger), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("heatmap_Hellinger.pdf")
heatmap_Hellinger
#dev.off()
#--------

# Geography Heatmaps --------
dat_loc_rel <- read.csv("rel_abund_location.csv")
dat_loc_rel
#Relative Abundance
heatmap_loc_rel <- ggplot(dat_loc_rel, aes(State, Clade)) +
  geom_tile(aes(fill = Rel_Abundance), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()

heatmap_loc_rel
#-------
dat_loc_rel$cube_root_rel <- (dat_loc_rel$Rel_Abundance)^(1/3) # make a new column with the square root of the relative abundance
heatmap_loc_rel_cube <- ggplot(dat_loc_rel, aes(State, Clade)) +
  geom_tile(aes(fill = cube_root_rel), colour = "white") +
  scale_fill_gradient(low = "white", high = "black") + theme_bw()
#pdf("heatmap_location_cube_root.pdf")
heatmap_loc_rel_cube
#dev.off()

```

##Metadata at tips of the Phylogeny
```{r}
# Location---
#Plot the tree with dots in various colors representing the broad location
plot_tree(physeq_prop, color="location_broad", base.spacing = 0.03, text.size=1)


#Level 3 Ontology----
#Plot the tree with dots in various colors representing the level 3 ontology category for the host
plot_tree(physeq_prop, color="level3", shape="level3", base.spacing = 0.1)
?plot_tree

#Using ggtree with a scale bar
#pdf("treelevel3.pdf")
ggtree(physeq_prop, ladderize = FALSE ) +
    geom_tippoint(aes( x = x + hjust, color=level3, shape=level3), size=1, alpha=.75) +
    scale_color_brewer("level3", palette="Greys") +
    theme_tree2(legend.position='right') 
#dev.off()

#pdf("treelevel3_color.pdf")
ggtree(physeq_prop, ladderize = FALSE ) +
    geom_point2(aes( x = x + hjust, color=level3, shape=level3), size=1, alpha=.75) +
    theme_tree2(legend.position='right') 
#dev.off()

#pdf("treeloc_color.pdf")
ggtree(physeq_prop, ladderize = FALSE ) +
    geom_point2(aes(x = x + hjust,  color=location_broad, shape=location_broad), size=1, alpha=.75) +
    theme_tree2(legend.position='right') 
#dev.off()

```

### Bar Chart
```{r}
# EMPO Level 3 Categories add up to 1-----
#pdf("bar_rel_abund_color.pdf")
ggplot(dat, aes(x = Level3, y=Relative.Abundance, fill = Clade)) + geom_bar(stat="identity")
#dev.off()

#pdf("bar_rel_abund.pdf")
ggplot(dat, aes(x = Level3, y=Relative.Abundance, fill = Clade)) + geom_bar(stat="identity") + scale_fill_grey()
#dev.off()

#Make sure phylogeny clade colors are the same as bar chart colors
hex_codes1 <- hue_pal()(7)  # Identify hex codes
hex_codes1  
show_col(hex_codes1)


# Clades add up to 1 ----- should be animal, plant, sediment, rock swab water
clade_rep <- read.csv("rel_representation_clades_inL3.csv")

clade_rep$Level3 <- factor(clade_rep$Level3, levels = c("Animal_corpus", "Plant_corpus", "Sediment", "Rock_swab", "Water"))

#pdf("rep_L3_per_clade_color.pdf")
ggplot(clade_rep, aes(x = Clade, y=Relative_Rep, fill = Level3)) + geom_bar(stat="identity")
#dev.off()

#pdf("rep_L3_per_clade_grey.pdf")
ggplot(clade_rep, aes(x = Clade, y=Relative_Rep, fill = Level3)) + geom_bar(stat="identity") + scale_fill_grey()
#dev.off()

#------
location_rel_State
location_rel_clade

```

### Host Co-Amplification
Look at samples with more and less than 50% host co-amplification and plot

```{r}
#Read in the old Phyloseq Object that included non-Exobasidiomycete reads ----
physeq <- readRDS("physeq1") #We are naming the phyloseq object we are working "physeq"
exops <- subset_taxa(physeq, Class=="Exobasidiomycetes")
tot_samps <- sample_sums(physeq) # the number of taxa found in each sample
length(tot_samps) #473
exosamps <- sample_sums(exops) #the number of class exobasidiomycete taxa found in each sample
length(exosamps) #473
host_ratio <- exosamps/tot_samps
sample_data(exops)$host_ratio <- host_ratio #add the host ratio as another column into the metadata
length(host_ratio) #473
success <- which(host_ratio > 0.5)
length(success) #211
```
